### 2021-03-02 math/ia_2021/paper.tex
ia work

% Suppresses headers, footers, and page numbers on title page
\begin{titlepage}
    \begin{center}
        \vspace*{4cm}
        A statistical approach to predicting a tsunami’s wave height \\
        \vspace{1cm}
        Can the height of a tsunami's wave be predicted based on the distance from and
        the magnitude of a preceding earthquake? \\
        \vspace{1cm}
        \textit{Väinö-verneri Kauppila} \\
        February 2021 \\
        \vspace{4cm}
        Word count: TBD
        \vfill
        \vspace{0.1cm}
    \end{center}
\end{titlepage}

@@

\subsection{Processing the dataset for use}
%TODO: #2 Ask for help 
The dataset, procured from the NOAA website \cite{noaa}, includes many tsunami
runups, going back before the year 0. Due to this, the NOAA states uncertainties
that can occur in its data. For example, reports that were recorded before the
20th century tend to be based on written accounts, and not recorded seismic activity.
Only since the 1960s, the installation of the Worldwide Standardized Seismograph
Network system allowed for more accurate data in this paper. For that reason, any recorded
activity before this date was discarded from the dataset for this paper. The dataset itself can
be accessed through two different methods; searching the database using the online
tool or downloading it in a somewhat obscure compressed Google Earth \verb|.kmz| format. 
The online tool supports the download of search results in a much more human-readable 
\verb|.tsv| (tab-separated) format. For this paper, all entries after the year 1900 were searched and downloaded.

@@

The data was split into a training and a test dataset. The training
dataset was used to generate a regression model while the test dataset was used 
to calculate the general accuracy of the model, by predicting values and
comparing those to the known value as found in the dataset. The dataset was split based on
a fraction using the \verb|df.sample| method found in the pandas library. A split
of 70\% for training and 30\% for test was chosen. A \verb|random_state| argument is
set to an arbitrary value 42, as if it weren't, the dataset would be split in a different
manner at each execution of the program.

@@

@@

where the term $\beta_0$ represents the y-intercept of the line and $\beta$ represents the
coefficient of $x$. The final term $e$ is what is known as the error or residual
term. It is an estimate of the amount by which the predictions differ from the real
value.

@@

Even though the right side of this equation is quadratic in $x$, this is still
linear regression as the right side is linear in $\beta$ (if we were to let
$x = 1$, we would be left with $y = a + \beta_1 + \beta_2 + e$, which is linear).
A second degree polynomial regression would best fit our model for both independent
variables, as seen in the graphs below, where a theoretical curve of regression is drawn
by hand:

@@

$SS_{res} = \sum_{i} (y_i - p_i)^2 = \sum_{i} e_i^2$, \footnote{$\sum_i i$ would mean that all values i 
are summed} where $e_i$ is the error term, equal to the difference of a data point $y_i$ 
and its corresponding predicted value $p_i$.

@@

\appendix
\section{Appendix}
\label{app}
\subsection{Python program}
\label{app:scripts}

>>>>

% Suppresses headers, footers, and page numbers on title page
\begin{titlepage}
    \begin{center}
        \vspace*{4cm}
        A statistical approach to predicting a tsunami’s wave height \\
        \vspace{1cm}
        Can the height of a tsunami's wave be predicted based on the distance from and
        the magnitude of a preceding earthquake? \\
        \vspace{1cm}
        \textit{Väinö-verneri Kauppila} \\
        February 2021 \\
        \vspace{4cm}
        Word count: TBD \\
        \textit{Note: diagrams reserve a large amount of space, and cannot be placed in 
        the appendix for the sake of clarity}
        \vfill
        \vspace{0.1cm}
    \end{center}
\end{titlepage}

@@

\subsection{Processing the dataset for use}
%TODO: #2 Ask for help 
The dataset, procured from the National Oceanic and Atmospheric Administration website \cite{noaa}, includes many tsunami
runups, going back before the year 0. Due to this, the NOAA states uncertainties
that can occur in its data. For example, reports that were recorded before the
20th century tend to be based on written accounts, and not recorded seismic activity.
Only since the 1960s, the installation of the Worldwide Standardized Seismograph
Network system allowed for more accurate data in this paper. For that reason, any recorded
activity before this date was discarded from the dataset for this paper. The dataset itself can
be accessed through two different methods; searching the database using the online
tool or downloading it in a somewhat obscure compressed Google Earth \verb|.kmz| format.
The online tool supports the download of search results in a much more human-readable
\verb|.tsv| (tab-separated) format. For this paper, all entries after the year 1900 were searched and downloaded.

@@

The data was split into a training and a test dataset. The training
dataset was used to generate a regression model while the test dataset was used
to calculate the general accuracy of the model, by predicting values and
comparing those to the known value as found in the dataset. The dataset was split based on
a fraction using the \verb|df.sample| method found in the pandas library. A split
of 70\% for training and 30\% for test was chosen. A \verb|random_state| argument is
set to an arbitrary value 42, as if it weren't, the dataset would be split in a different
manner at each execution of the program. The program can be found in Appendix \ref{app:scripts}.

@@

\subsubsection{Linear model}

@@

where the term $\beta_0$ represents the y-intercept of the line and $\beta$ represents the
coefficient of $x$. The final term $e$ is what is known as the error or residual
term. It is an estimate of the amount by which the predictions differ from the real
value. An example of linear regression can be found in Appendix \ref{app:regression}, where 
a set of data points is presented on a scatter plot, and the regression line is found 
by finding the individual coefficients $\beta_n$ of $x$ (see equation above), like such:

$\hat \beta_1 = \frac{\sum (x_i - \bar x) (y_i - \bar y)}{\sum (x_i - \bar x)^2}$, where 
$\hat \beta_n$ represents an estimated value and $\bar x$ represents the mean of the set $x$

and the $\hat \beta_0$ value can be approximated as $\hat \beta_0 = \bar y - \hat \beta_1 \bar x$

For the dataset in \ref{app:regression}, the results to these equations are:

$\hat \beta_1 = \frac{1806757.14}{322280.86} = 5.60$ and $\hat \beta_0 = 2257.14 - 5.61 \cdot 512.86 = -618.01$

So the regression line can be expressed as $\hat y = 5.60 x -618.01$

\subsubsection{Special cases}

@@

Even though the right side of this equation is quadratic in $x$, this is still
linear regression as the right side is linear in the $\beta$ terms (if we were to let
$x = 1$, we would be left with $y = a + \beta_1 + \beta_2 + e$, which is linear).
A second degree polynomial regression would best fit our model for both independent
variables, as seen in the graphs below, where a theoretical curve of regression is drawn
by hand:

@@

$SS_{res} = \sum_{i} e_i^2 = \sum_{i} (y_i - p_i)^2$, \footnote{$\sum_{i=1}^n x_i$ means that all values of
    x are summed, beginning from $x_1$ and ending at $x_n$} where $e_i$ is the error term, equal to the difference of a data point $y_i$
and its corresponding predicted value $p_i$.

@@

\appendix
\section{Appendix}
\label{app}
\subsection{Example linear regression}
\label{app:regression}

Below is an example of a dataset with $x$ values and their corresponding
Y values, the data points chosen are purely arbitrary.

\begin{center}
    \begin{tabular}{||c c||}
        \hline
        X values & Y values \\ [0.5ex]
        \hline\hline
        368      & 1700     \\
        \hline
        340      & 1500     \\
        \hline
        665      & 2800     \\
        \hline
        954      & 5000     \\
        \hline
        331      & 1300     \\
        \hline
        556      & 2200     \\
        \hline
        376      & 1300     \\ [1ex]
        \hline
    \end{tabular}
\end{center}

\pgfplotstableread{
    X Y
    368 1700
    340 1500
    665 2800
    954 5000
    331 1300
    556 2200
    376 1300
}\table

Below is the dataset plotted, along with a line of regression.

\begin{tikzpicture}[scale=1.1]
    \begin{axis}[
            title={Comparing X and Y},
            xlabel={X axis},
            ylabel={Y axis},
            xmin=300, xmax=1000,
            ymin=0, ymax=5500,
            ytick={0,1000,2000,3000,4000,5000},
            xtick={0,400,500,600,700,800,900},
            legend pos=north west,
            ymajorgrids=true,
            grid style=dashed,
            scale=1.2
        ]

        \addplot[
            only marks,
            color=blue,
            mark=*,
        ] table {\table};
        \legend{Datapoint}

        \addplot [thin, blue] table[
                y={create col/linear regression={y=Y}}
            ] % compute a linear regression from the input table
            {\table};
        \addlegendentry{
            Linear regression: $ x =
                \pgfmathprintnumber{\pgfplotstableregressiona}
                \cdot t
                \pgfmathprintnumber[print sign]{\pgfplotstableregressionb}$
        };

    \end{axis}
\end{tikzpicture}

\subsection{Python program}
\label{app:scripts}
